Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-11-13 paper/introduction.tex
fix

First, causal and non-causal factors may be numerous and collinear. In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or a change in pressure, as these two factors may, at first, be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factors increases. In neuroscience, for example, identifying whether the frequency of a word presented on a subject's retina modulates brain activity can be surprisingly difficult. Indeed, the frequency of words in natural language covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs) and so forth \citep{kutas2011thirty,pegado2014timing}. Instead of selecting a set of words that control for all of these factors simultaneously, it is thus common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, e.g. \citep{huth2016natural}), and investigate, in the model, the estimated contribution of each factor \citep{friston1994statistical}.

@@

Here, we introduce the `back-to-back regression' (B2B), which not only combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}), but also provide robust and interpretable causal coefficients. After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques in identifying causal influence (Section~\ref{sec:experiment_synthetic}). Finally, we apply B2B to a large neuroimaging dataset and reveal that distinct but collinear word features lead to distinguishable brain representations (Section~\ref{sec:experiment_real}).

>>>>

First, causal and non-causal factors may be numerous and collinear. %In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or a change in pressure, as these two factors may, at first, be difficult to manipulate independently.
This issue becomes increasingly pronounced as the number of potential factors increases. In neuroscience, for example, identifying whether the frequency of a word presented on a subject's retina modulates brain activity can be surprisingly difficult. Indeed, the frequency of words in natural language covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs) and so forth \citep{kutas2011thirty,pegado2014timing}. Instead of selecting a set of words that control for all of these factors simultaneously, it is thus common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, e.g. \citep{huth2016natural}), and investigate, in the model, the estimated contribution of each factor \citep{friston1994statistical}.

@@

Here, we introduce the `back-to-back regression' (B2B), which not only combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}), but also provide robust, interpretable and unidimensional causal coefficients for each tested factor.

The present paper focuses on the restricted issue of disentangling the causal influence of linearly correlated predictors ($X$) onto multivariate observations ($Y$). This issue thus differs from other causal discovery algorithms based on temporal-delays and/or nonlinear interaction in systems where the directionality of causation (from X to Y or vice versa) is unknown (e.g. \citep{peters2017elements, granger1969investigating, janzing2013quantifying, scholkopf2016modeling}.

After detailing B2B method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques in identifying causal influence (Section~\ref{sec:experiment_synthetic}). Finally, we apply B2B to a large neuroimaging dataset and reveal that distinct but collinear word features lead to distinguishable brain representations (Section~\ref{sec:experiment_real}).

### 2019-11-13 paper/paper.tex
fix

We perform two experiments to evaluate B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography (MEG) dataset.
%
We use scikit-learn's CCA, PLS, and RidgeCV objects to compare B2B against the standard baselines \citep{sklearn}.

@@

% I find a bit surprising to mention cauaslity literare in ML at the end. I would have expected this in the introduction to convince reviewer that know your stuff.

Finally, B2B is also related to, but different from causal discovery algorithms used outside the neuroimaging field. These methods are often nonlinear \citep{peters2017elements} and base their causal estimates on temporal delays \citep{granger1969investigating, janzing2013quantifying}. For example, B2B relates to the half-sibling regression \citep{scholkopf2016modeling}, in the sense that we observe a large number of variables corrupted by the same source of noise; we leave the extension of B2B to nonlinear time-series for future work.

>>>>

We perform two experiments to evaluate B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography (MEG) dataset.
%
We use scikit-learn's PLS and RidgeCV \citep{sklearn} as well as pyrcca' regularized canonical compnent analysis (RegCCA, \citep{bilenko2016pyrcca}) objects to compare B2B against the standard baselines.

@@

% I find a bit surprising to mention cauaslity literare in ML at the end. I would have expected this in the introduction to convince reviewer that know your stuff.
CONCLUSION SENTENCE
